Unclear Regional Effects of a Large Marine Protected Area Network

### 2020-04-19 documents/reviewer-responses.Rmd

---
title: "Reviewer Responses for 'Unclear Regional Effects of a Large Marine Protected Area Network' (previously 'The Regional Effects of MPAs')"
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
output:
  word_document: default
  html_document: default
  pdf_document: default
bibliography: My Library.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reviewer Responses

We thank the two reviewers and the Editor for their constructive responses. We have fundamentally restructured the paper to place much greater emphasis on the empirical results from the Channel Islands MPA network. This is to our knowledge the first empirical assessment of the regional effects of a large and iconic MPA network on a wide assemblage of commercially and economically important species. Our inability to estimate a clear effect in such a well-enforced and well-studied local is we feel a vitally important "null-result" both for scientists and MPA managers to be aware of. As such we feel that it warrants publication in PNAS.

While our results are more focused on the empirical results, we explicitly link our results to the large body of MPA theoretical work, using it to help illustrate why we might expect such a null result and help managers understand when they might be able to detect an effect. As MPAs are used as a core tool of marine resource management, our results provide a critical update to our empirical expectations for MPAs, that to date have been based almost entirely on inside-outside MPA response ratios. We hope you find the updated version of our manuscript improved thanks to the inclusions of the careful points raised by the prior revision of this paper.

<!-- In addition to the comments from the last round, we have made a change in the estimation model that we feel should be highlighted. The previous draft of this manuscript utilized a mixed-effect model to estimate the regional effects of MPAs, fitting to data at the level of individual transect observations. This model provides a robust accounting of uncertainty at multiple levels, but it also cumbersome and   -->

## Reviewer 1

- *I think that greater emphasis needs to be placed up front, abstract, statement of significance, introduction..perhaps even title, that this model and paper only addresses fish biomass...not other 'conservation outcomes'. You do define this but not until page 2, line 148-154. I recommend moving this up.*
  - We apologize for not making this clearer in our earlier draft, and have included this key caveat in the abstract, statement of significance, and introduction (Line 98)

- *I was surprised that there was no discussion about enforcement being a potential confounding factor. Currently the paper assumes that the Channel Islands MPAs are fully enforced, without poaching. This may or may not be the case and should be stated as such. You do address well other system wide impacts like water temperature (lines 557-8, 691+). Perhaps enforcement could be added to the discussion around line 711-13.*
  - This is indeed a critical point that we have endeavored to correct. We have included enforcement under our summary of key enabling conditions for effective MPAs (line 124), as well as the discussion (line 382). While we do not have published data on strength of enforcement within the MPAs, anecdotal evidence suggests it is relatively high, as well as evidence from the mainland California MPAs [@zellmer2018] indicates it is high. In addition, the presence of a clear increase in response ratios within the Channel Islands provides evidence that the MPAs  are large enough and sufficiently well-enforced as to provide meaningful protection within their borders (Line XX).

- *Your discussion of the model also emphasizes that stock status prior to MPA designation is very key to outcomes; in the SI most of the species included for the Channel Island review have unknown status---which influences your discussion, and may better be described.*
  - We make clearer that we iterate over unknowns such as pre-MPA stock status in our description of the Channel Islands simulations (line XX), and the implications of pre-MPA stock status in the discussion (Line XX).

## Reviewer 2

- *Yes, these are important questions but they are not new to ecology, economics, and fishery science. You are not the first to ask them and point out their importance; see articles at bottom for some examples. What are you doing to answer them that is better than all of the other papers that come before you? And why is your case study an exemplary place to answer the questions? The conclusions on line 709-728 are well-known in the MPA literature.*
  - We have fundamentally restructured the paper based around the importance of the Channel Islands as a case study in measuring MPA effectiveness. We thank the reviewer for pointing out gaps in our literature, and we have incorporated the studies they recommend. The theoretical points we raise have indeed been pointed out in the literature before, and we feel our new manuscript makes that clearer. The Channel Islands MPAs are an iconic network that has served as a model for protected areas around the world. It is also incredibly well-studied and well-enforced. While previous studies have looked the effectiveness of the Channel Islands MPAs within their borders, this is the first study in the Islands (and globally of which we are aware) to attempt an estimate of the effect of the MPAs at the regional scale (both inside and outside of the MPAs themselves). That we are unable to estimate a clear regional effect in even this system is an important result that we feel warrants publication in PNAS. As MPAs are increasingly turned to tools for ocean conservation, our results provide an important update and caution to expectations based solely on inside-outside comparison studies.

- *I would also be more cautious regarding your conclusion about your model predicts what should be expected in the data analysis. It could be simply luck that there is agreement between the two.*
  - We have made the language clearer that this concordance could be simple coincidence, and is not in any way proof of the validity of our results (Line XX) However, in light of recent concerns in the scientific literature around hypothesizing-after-results, we feel that it is critically important to ground our theoretical results in *a priori* predictions. In addition, as we mention in the discussion, we encourage MPA scientists and stakeholders to undertake such simulation modeling to consider the plausibility of their results. As the reviewer says, empirical results that match theoretical predictions are not proof that the empirical methods are robust, but they do provide am additional and important line of evidence: empirical results that fail to match theoretical predictions need to be carefully considered.

- *What about using the simulation model to demonstrate the validity of the technique used to identify the MPA effect used in the paper, which was pioneered in publication in 38? Or to understand whether the implication of the statement on line 539, that is, species ranges exceed the Island and your sampling domain, is important for measuring the impacts of MPAs outside of their boundaries? In sum, in its current form, I would move all of the simulation model to the SI (lines 131-252) and mention in the discussion section that these results are consistent with the predictions of a stylized single species model.*
    - We have greatly expanded the simulation validation of our method, which was previously included in the supplementary material. Figure 4 of the paper now demonstrates the validity (and shortcomings) of the model under a variety of plausible conditions. We would note that this method is related to that described in @caselle2015 (previously reference #38), but is not the same. Our method explicitly estimates the net difference in the targeted and non-targeted biomass densities pre-and-post MPA as an estimate of the mean effect of the MPAs  both inside and outside their borders. Caselle et al. 2015 used trends in targeted and non-targeted species inside and outside of MPAs to illustrate the effects of the MPAs within their borders. We stress this difference in an effort to make clear that our results and methods are not simply an update to Caselle et al. 2015, but address a fundamentally different question. We hope we have made this clearer in the text, e.g. on line XX.

    With regards to the specific question about the possibility of the species range exceeding the sampling domain, this would be reflected as a reduction in the percent of the population protected by the MPAs: a 25km^2 MPA may provide near total protection for a sedentary species with brood-rearing, but be insignificant to a migratory tuna. To that end, our assessment of model performance as a function of MPA effect size addresses the question of how MPA size relative to species range affects our models performance. All else being equal the smaller a portion of a population's range is protected by MPAs, the smaller the expected effect size, and by extension the more challenge the estimation model will have.

  We apologize that we did not present our prior simulation work in a sufficiently compelling manner, and failed to differentiate its results from the prior literature. As suggested by the reviewers and the editor, we have restructured the paper to place much greater emphasis on the critical results from the Channel Islands. However, we feel that the simulation results do provide vital context to our results, and provide tangible recommendations as to under what circumstances we might be able to estimate the regional effects of MPAs. The reviewer correctly points out that the individual components of our model are not themselves novel, and that the importance of fleet dynamics and fish dispersal have been previously discussed. As such as have greatly reduced the attention these results receive, except as a vehicle for the explicit consideration of the feasibility of estimating regional MPA effects. In addition, while these broad points have been made in the literature before, we feel there is value in cohesively demonstrating the ramifications of these diverse pieces of MPA theory (that generally are looked at only one or a few at a time) in once place, both through our Figure 5 and the accompanying web application. We hope that these modifications make clear the contribution of the simulation modeling.

- *The discussion on how to detect changes outside the MPA (lines 254-347) follows very closely the discussion in Ferraro et al (#41 in your list). It seems like you can remove most of this text and point the reader to that paper*
  - We have greatly reduced the amount of text addressing response ratios, and restructured the response-ratio analysis to be linked explicitly with the empirical response ratios observed in the Channel Islands (in response to comments from the editor), in Figure 1. We feel this strategy makes much clearer how the results of this paper expand beyond the findings of Caselle et al. 2015, and why the results of Caselle et al. 2015 are evidence of within-MPA effects, but not of the question of regional effects addressed in this manuscript. Given that much of the MPA literature relies on response ratios as the empirical evidence for the effectiveness of MPAs, we feel that it is critically important to directly address the challenges of response ratios as measures of regional effects in this paper. We hope that the new manner, directly linking theoretical response ratios to the levels observed in the Channel Islands.

- *To me the potential contribution of the paper starts on line 350, which is way too late in a paper of this size*
  - We have restructured the paper so that it now focuses on the Channel Islands case study, and as suggested have moved this discussion up to the introduction (line 97).

- *Is part of assumption b) (line 372) that the dispersal characteristics of the fished and non-fished species need to be similar? I could envision a scenario where the two sets of species have different dispersal characteristics, non-fished have smaller dispersal ranges than fished species, for example. How does this impact your identification strategy? If the trends are similar regardless of the mechanistic process by which it occurs, does it matter for the empirics? More discussion on both assumptions (a) and (b) and what they imply about the biological mechanisms seems warranted.*
  - The estimation model does not depend on the dispersal characteristics of the fished and non-fished species per-say. The model is operating at the level of biomass densities. As such, even if a non-targeted species has a much bigger range than the targeted species (or even the sampling region), so long as the parallel trends assumption holds within the sampled portion of the non-targeted species range, the identification strategy may be valid. Similarly, the model is agnostic as to the root causes of the parallel trends, so long as they exist. However, if for example the parallel trends are dependent on a particular temperature regime which during the study is disrupted (as is likely the case in our data), then the model may be reliable for the early years but not later. We have made this point clearer in the paper (line XX), and made clear that the lack of a clear effect in the recent years is likely due to an emerging violation of the parallel trends assumption, and not to a failure of the MPAs per say. However, this is still an important result, since it suggests that neither the response ratios nor the difference-in-difference method give us a robust estimate of the regional effects of the Channel Island MPAs. As we stress in the paper, that even in a network such as the Channel Islands we are left without a clear empirical effect at the regional level has critical ramifications for the design and monitoring of MPAs globally.

- *The use of visual data surveys has, of course, problems that introduce errors in dependent and independent variables that could be correlated with features of the fished and non-fished species, e.g., size distribution. How can you be sure that this is not driving your results? Could you use the simulation model to investigate this?*
  - Visual survey data are indeed challenging, but also often the only form of data available for MPA analysis. Our base model operates at the level of total biomass density of targeted and non-targeted species, at the site level. This is the finest spatial scale at which the sampling regime would be likely to provide an estimate of the variable in question, since while any individual transect is unlikely to observe all of the targeted / non-targeted, but aggregating across transects provides a better chance of this. These totals are derived from average biomass densities of individual species across multiple transects within each site, which are then summed across species to achieve the total mean biomass density at the site level. The averaging process in acts to smooth out some of the sampling variation, and reflects a broader discussion about the right degree of smoothing in spatio-temporal data.

How might this affect our results? So long as sampling errors are consistent over time, they should not bias the mean or the trends of our results (e.g. if we systemically under/over estimate the biomass of some species). There is no doubt though that we are underestimating the true magnitude of the credible intervals though. For example, the model assumes constant length-to-weight relationships over time, whereas in reality there is clearly variation both among individuals and over time. Accounting for this should add more noise to our data, further complicating the ability of our model to achieve precise estimates. Complicated relationships such as the MPAs themselves affecting the length-to-weight relationship would indeed skew our results, but we feel that is beyond the scope of this or any other MPA study of which we are aware to address.

  We have run our model at a wide array of resolutions, from the updated aggregated form we use here to a complete hierarchical framework that attempts to control for errors in the individual observation process (Fig.SXX). The fundamental results remains the same: a general upwards trend, followed by a downward trend, with the clearest signal in the years around 2010. We would actually be more concerned about the robustness of our results if we had found a precise but small effect size: such an effect could easily disappear if we accounted for errors in the data (e.g. through allowing for variations in the length-weight relationship). Our relatively uncertain findings instead suggests that it is unlikely that we would get fundamentally different conclusions by including more error; we would simply be even less certain than the uncertain state the model estimates we are in. We have expanded the simulation testing of the method a bit to help illustrate how violations of the model assumptions play out, and how they are related to the magnitude of the effect size (Figure 5). The full suite of alternative resolution models can be found in Figs.SXX-XX (all of which show the same general result), and a a simulation test of our model incorporating factors such as differing diver skill can be found in Fig.SXX (which shows that the model is capable of performing under such circumstances).

- *(The answer to your question on line 709 that appears on lines 714-719 has been known in the literature on MPAs for over 20 years; see, e.g., papers by Martin Smith below. Given that your question is what "your results imply about the future of marine science", my answer is not very much. Again, you did a rigorous empirical case study that should be published. I am however finding it difficult to see a path to publication in PNAS.)*
  - We hope that our restructuring around the novel empirical results resolves this concern. While we agree that framing our results around the future of marine science would be a step too far, we would point out that our earlier and present draft asked what our results imply for the future of MPA science more specifically, which we hope is less of a stretch.

  Given the increasing role of MPAs in conservation strategies, the lack of a clear regional effect in a place like the Channel Islands should be known and taken into account by MPA practitioners around the globe. We agree that we framed our simulation results poorly in previous drafts, but feel that they do compliment the empirical results to help managers understand why the Channel Islands results may not be surprising, and when they might expect to be able to detect and effect, and when not. See section 2.1.2 ("When can we detect the effects of MPAs") and our updated conclusions, in particular the paragraph beginning on line XX.

- *Examples of relevant literature, you should expand your net to find other relevant literature to better distinguish your contribution*
  - We thank the reviewer for pointing out these important omissions. We have included each of these references and their key messages in their appropriate portions of the paper.

# References

<!-- ## Editor -->

<!-- # Original Reviews  -->

<!-- Dear Dr. Ovando, -->

<!-- I regret to inform you that the PNAS Editorial Board has rejected your manuscript [MS# 2019-17437]. The expert who served as editor obtained 2 reviews, which are included below. After careful consideration, the editor decided that we cannot accept your manuscript. -->

<!-- However, because the reviewers think the work is of interest and the editor concurs, we are willing to consider a re-submission that constructively addresses all of the concerns raised in the critiques. The paper would have to satisfy both the reviewers and the editor, and new criticisms could arise upon re-evaluation. We cannot guarantee success and will be unable to consider further re-submissions. -->

<!-- Please submit the re-submission as you would a new paper. Your re-submission will be given a new tracking number. In your cover letter and the Comments to Editorial Staff, please specify that the paper is a re-submission and include your original tracking number. Additionally, please provide a "re-submission/revision" cover letter, including a point-by-point response to the editor and reviewers' comments. If you have any questions, please contact our office. -->

<!-- Thank you for submitting your work to PNAS. -->

<!-- Sincerely yours, -->
<!-- May R. Berenbaum -->
<!-- Editor-in-Chief -->

<!-- Editor's Remarks to Author: -->

<!-- This is a fascinating manuscript because it examines an iconic MPA, for which there may be more detailed biological and physical data than any other MPA. -->

<!-- The simulation model that is applied to the case study is also rich in detail. -->

<!-- One referee, who is a world-renowned fisheries scientist with quantitative chops and deep knowledge, challenges the novelty and significance of the paper, holding it to a "high PNAS-bar", while also recognizing that as a case study it needs to be published. I hold this referee in high regard, and upon a close reading of the manuscript myself, I had to agree that when I was leading a science group at NOAA fifteen years ago, which is when MPA's were a hot new topic, it was widely appreciated that the outcome of any MPA would be dramatically impacted by where displaced vessels go and movement rates of adult fish. -->

<!-- In order for this to be published in PNAS, you need to be very clear on the manuscript's novel contribution, and also place it better in the salient literature. How relevant are references 33 and 35 compared to the references provided by referee #2? -->

<!-- The one difference I have with referee #2 is that I am not sure to what extent the richness or many, many different species, of before and after data you have for the Channel Islands MPA has ever been matched in other MPA study? In other words, there is some chance the empirical data may be what are most novel - although those data do not come to the forefront in this paper. -->

<!-- It could be that this warrants two papers rather than one. I would have liked to seen a paper focusing on the results of the difference in differences analyses. For maximum impact these would be presented in a naïve form because that is illustrative to practitioners-as it is "what they expect", and in the more rigorous form after accounting for the absence of any true control site and modeling the fact that an MPA affects the so-called, but non-existent, control. The naïve results are worthy of attention. If the empirical data were elevated, then the Supplementary materials need to include more information on the sampling, spatial and temporal detail, etc. As it is, I have no idea how much confidence to put in the fish-by-fish biomass data. -->

<!-- Then the great merit of the model is it should tell the manager what they should expect, and what they need to do if they want to make sure any newly instituted MPA is performing as hoped for. There are several marine conservation NGO's who are operating with totally naïve and misguided expectations about what will be evident after an MPA is established. Written up well, this research could mitigate some dangerous delusions. To do this the model would be used to discuss: -->

<!-- What would you expect to see even if an MPA was working well? -->

<!-- What would you expect to see if it an MPA was ineffective and doomed to be ineffective? -->

<!-- What does all this mean for the expectations we put on MPA's? -->

<!-- All of the above questions are touched on in this manuscript -- but not direct enough for a manager or NGO to take note of them. Lines 642 to 651 are noteworthy, but are kind of lost in their placement. -->

<!-- Reviewer Comments: -->
<!-- Reviewer #1: -->

<!-- Suitable Quality?: Yes -->
<!-- Sufficient General Interest?: Yes -->
<!-- Conclusions Justified?: Yes -->
<!-- Clearly Written?: Yes -->
<!-- Procedures Described?: Yes -->
<!-- Supplemental Material Warranted?: Yes -->

<!-- Comments on Significance Statement: -->

<!-- I think it could use some clarification/specificity. Since fish biomass was the only 'conservation outcome' analyzed I think that should be stated in the significance statement as well as abstract. Further I think an additional word in the case study sentence would be helpful to demonstrate that was the empirical part of the study, additional words in caps: "We find no statistically clear effect of ONE EXISTING network of MPAs in FISH BIOMASS after 13 years of protection, which our simulation analysis suggests is actually to be expected." -->

<!-- Comments: -->

### 2020-04-19 documents/ovando-regional-effects-of-mpas-v2.Rmd
version sent to coauthors

---
title: Regional Effects of Marine Protected Area Networks Are Unclear
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world's oceans, up from less than 1\% in the year 2000. The Convention on Biological Diversity called for 10% of coastal waters to be protected inside MPAs by 2020, while the International Union for Conservation of Nature is now calling for 30\% protection by 2030. It is often clear that MPAs produce conservation benefits inside their borders, but many MPAs are also justified on the grounds that they may benefit the broader region outside their borders. The conservation effects of MPAs are most commonly evaluated using response ratios---measures of numerical or biomass densities inside their borders, relative to these metrics in reference sites outside their borders. Studies of this nature have provided broad evidence that MPAs produce conservation benefits within their borders. While these gains can be vitally important, marine populations are rarely contained entirely within the borders of MPAs; therefore a critical question is not only do MPAs produce conservation gains inside their borders, but how do they affect the broader region in which they are located?  The Channel Islands National Marine Sanctuary provides a clear example of the importance and challenges of estimating regional MPA effects. A network of MPAs covering 20\% of the Islands’ waters was put in place in 2003, with a goal of providing regional conservation and fishery benefits. Response ratios from the region indicate that the Channel Island MPAs have increased biomass densities inside the MPAs. We are unable to find a clear effect of these same MPAs on biomass densities throughout the Channel Islands region. We use a bio-economic simulation model to explain this discrepancy, and demonstrate under what conditions we likely can and cannot clearly estimate the regional conservation effects of MPAs. MPA networks covering 25\% or less of a region are likely to produce regional increases in fish biomass densities on the order of 25\% or less, a meaningful effect but also one that can easily be overwhelmed by environmental shocks. Our results provide a novel assessment of the regional effects of a large and iconic Marine Protected Area network, and provide guidance for managers charged with monitoring and adapting MPAs. 
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::pdf_document2: default
  bookdown::word_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
---

@@

## significance
  Healthy marine ecosystems are critical to the well-being of the planet. Marine protected areas (MPAs), parts of the oceans protected from human activities such as fishing, are increasingly being used to conserve and manage these ecosystems. Our study pairs theory with empirical methods to show that the regional conservation effects of marine protected areas can be highly variable and dependent on human drivers. We find no statistically clear effect of a large and well protected network of MPAs in the Channel Island, California, after 14 years of protection, which our simulation analysis suggests is actually to be expected. MPA practitioners must clearly communicate expectations from MPA networks, and plan monitoring and adaptations strategies that directly address the challenges of estimating regional conservation effects of protected areas.

## acknowledgements
  Funding for the simulation model was provided by the NMFS-Sea Grant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not be possible without the work provided by PISCO divers over the years, with special thanks to to Kathyrn Davis Koehn and Avrey Parsons-Field. DO thanks Cody Szuwalski, Julia Lawson, and André Punt for helpful comments and technical support

@@

Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], larval dispersal patterns and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess2014), the age and degree of enforcement [@edgar2014],  how overfished the population would be without the MPA, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006;@walters2000a;@white2011]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through network effects [@gaines2010; @costello2010a].

@@

The MPAs in this study cover 21% of the the surface waters of the Northern Channel Islands, a region spanning roughly 90km east to west. Detailed dispersal studies are not available for all of the species covered by our study, but what information we have suggest that while adults of some of these species are likely to exhibit site fidelity [@love2011], larvae are likely distributed beyond the Channel Islands [@rassweiler2012]. Therefore, if we assume that the population of the targeted species is at least equal to the extent of the Channel Islands, we can assume that at most the MPAs cover roughly 20% of the targeted populations' ranges.

Turning to the critical question of fishing pressure, formal stock assessments are largely lacking for these species. However, what evidence we have suggests mostly moderate fishing pressure, with some species such as California sheephead (*Semicossyphus pulcher*) and blue rockfish (*Sebastes mystinus*) experiencing high levels of fishing mortality / biomass  levels below target levels during the early 2000s [@alonzo2004;@dick2017].

```{r}
median(sim_mpa_effects$mpa_effect[sim_mpa_effects$years_protected> 13])
```

We integrated these network design features, life history traits, and exploitation histories into a spatially explicit bio-economic simulation model, and used this model to generate expected outcomes for the Channel Islands MPAs. Our results suggest that while a wide range of outcomes are plausible, from 0% to upwards of 200%, the a median simulated effect size for a stylized version of the Channel Islands after over ten years of protection was roughly `r percent(median(sim_mpa_effects$mpa_effect[sim_mpa_effects$years_protected > 10]))`; a potentially small value from the perspective of empirical detectability given the natural variability and sampling challenges of marine environments.

@@

Having established a theoretical grounding for the likely biomass density outcomes of the Channel Island MPAs, what empirical evidence do we have of the effects? @caselle2015 examined changes in biomass densities of species targeted by fishing activity inside and outside of the MPAs over time (a metric generally termed a "response ratio"), and compared these changes to the trends in non-targeted species. They found a statistically significant increase in the response ratios of targeted species over time, and evidence that this increase is smaller in the non-targeted species. Updating the results of @caselle2015 through 2017 with a Bayesian response-ratio regression shows a continuation of the increasing trend in the response ratios of targeted species (Fig.\@ref(fig:rr-plot)).

Why do these response ratio results not serve as sufficient evidence for the regional effects of the Channel Island MPAs? Control sites used in calculating response ratios are often selected based on abiotic or ecological traits such as habitat characteristics [@ferraro2018]. However, selection of control sites is further complicated by the very spillover that MPAs are often intended to create. Export of adults or larvae from the MPA to the "control" site affects their status as controls, as does displacement of fishing effort from MPAs to control sites. In theory, control sites far enough away to negate both biological spillover and concentration by the fishing fleet could be selected, but finding suitably far sites that are also appropriate proxies for the ecological and economic context of the MPAs may be challenging. While these concerns have been stated previously [e.g. @halpern2004;@white2011], the MPA evaluation literature has by and large been unable to adequately address them, often due to the very real logistical challenges of identifying and sampling adequate control sites that are truly independent of the MPAs in question. As a result of these challenges, response ratios can be a highly imprecise and biased measure of the regional conservation effect of an MPA network [@ferraro2018].

@@

While we do not have access to fine scale fishing data from the Channel Islands alone, reported catches for the species of interest in the Santa Barbara region in fact exhibit an overall downward trend in the years post reserve (see Fig.SXX). We can most likely rule out a negative MPA effect caused by a constant-catch fishing strategy then.  What is another possible explanation for the recent downward trend in the estimated MPA effects? The Channel Islands region (and the entire West coast of the USA) experienced a dramatic ‘marine heatwave’ beginning in 2014 and persisting through 2016, resulting in part in extremely elevated water temperatures throughout the region  [@gentemann2017]. Biogeographic differences in the distributions of targeted and non-targeted species may confound the observed effects of MPAs. Many of the non-targeted species in the Channel Islands have warm thermal affinities and have increased in numbers since the heatwave [@freedman2019]. The targeted group is made up mostly of fishes with cold-water affinities, such as members of the genus *Sebastes*. We hypothesize that the recent evidence for a decline in densities of targeted species is due to environmental conditions that disproportionately affect the targeted group (and not for example due to concentrated fishing pressure outside the reserves). This hypothesis is supported by the declining trend in mean biomass densities of targeted species seen inside the MPAs themselves. If the cause of recent declines was due to increases in fishing pressure we would expect to see substantial declines only in the fished areas (Fig.\@ref(fig:raw-density)).

>>>>

---
title: Unclear Regional Effects of a Large Marine Protected Area Network
date: "`r Sys.Date()`"
author: 
  - Daniel Ovando
  - Jennifer E. Caselle
  - Christopher Costello
  - Olivier Deschenes
  - Steven D. Gaines
  - Ray Hilborn
  - Owen Liu
abstract: |
   Marine Protected Areas (MPAs) cover 3-7% of the world's oceans, up from less than 1\% in the year 2000. The Convention on Biological Diversity called for 10% of coastal waters to be protected inside MPAs by 2020, while the International Union for Conservation of Nature is now calling for 30\% protection by 2030. It is often clear that MPAs produce conservation benefits inside their borders, but many MPAs are also justified on the grounds that they may benefit the broader region outside their borders. The conservation effects of MPAs are most commonly evaluated using response ratios---measures of numerical or biomass densities inside their borders, relative to these metrics in reference sites outside their borders. Studies of this nature have provided broad evidence that MPAs produce conservation benefits within their borders. While these gains can be vitally important, marine populations are rarely contained entirely within the borders of MPAs; therefore a critical question is not only do MPAs produce conservation gains inside their borders, but how do they affect the broader region in which they are located?  The Channel Islands National Marine Sanctuary provides a clear example of the importance and challenges of estimating regional MPA effects. A network of MPAs covering 20\% of the Islands’ waters was put in place in 2003, with a goal of providing regional conservation and fishery benefits. Response ratios from the region indicate that the Channel Island MPAs have increased biomass densities inside the MPAs. We are unable to find a clear effect of these same MPAs on biomass densities of targeted finfish throughout the Channel Islands region. We use a bio-economic simulation model to explain this discrepancy, and demonstrate under what conditions we likely can and cannot clearly estimate the regional conservation effects of MPAs. MPA networks covering 25\% or less of a region are likely to produce regional increases in fish biomass densities on the order of 25\% or less, a meaningful effect but also one that can easily be overwhelmed by environmental shocks. Our results provide a novel assessment of the regional effects of a large and iconic Marine Protected Area network, and provide guidance for managers charged with monitoring and adapting MPAs. 
bibliography: My Library.bib
csl: pnas.csl
output:
  bookdown::word_document2: default
  bookdown::pdf_document2: default
  bookdown::html_document2: default
params:
  run_name: ["v6.0"]
  sim_years: [50]
  burn_years: [20]
  num_patches: [25]
linkcolor: blue
toc: false
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}\linenumbers
  - \usepackage{draftwatermark}
---

@@

## significance
  Healthy marine ecosystems are critical to the well-being of the planet. Marine protected areas (MPAs), parts of the oceans protected from human activities such as fishing, are increasingly being used to conserve and manage these ecosystems. Our study pairs theory with empirical methods to show that the regional conservation effects of marine protected areas can be highly variable and dependent on human drivers. We find no statistically clear effect of a large and well protected network of MPAs in the Channel Island, California, on targeted finfish after 14 years of protection, which our simulation analysis suggests is actually to be expected. MPA practitioners must clearly communicate expectations from MPA networks, and plan monitoring and adaptations strategies that directly address the challenges of estimating regional conservation effects of protected areas.

## acknowledgements
  Funding for the simulation model was provided by the NMFS-Sea Grant Population and Ecosystem Dynamics Fellowship. Empirical data collection was funded primarily by the David and Lucille Packard Foundation in support of the Partnership for Interdisciplinary Studies of Coastal Oceans (PISCO) with additional funding from the California Ocean Protection Council and California SeaGrant. This study would not be possible without the work provided by PISCO divers over the years, with special thanks to to Kathyrn Davis Koehn and Avrey Parsons-Field. We also thank Cody Szuwalski, Julia Lawson, André Punt, and two anonymous reviewers for helpful comments and technical support.

@@

Numerous factors can affect the regional effects of MPAs. These include the scale of adult and larval dispersal relative to the size of the MPAs  [@gaines2003; @botsford2008; @difranco2018;@mcgilliard2008;@gerber2005], larval dispersal patterns and the strength and timing of density dependence in the population (e.g. pre- or post-settlement, @burgess2014), the age and degree of enforcement [@edgar2014],  how overfished the population would be without the MPA, and how fishing and management responds to the implementation of the MPAs [@hilborn2004; @hilborn1992; @walters2004; @hastings2003;@gerber2003; @hilborn2004a; @gaines2010; @botsford2003;@smith2003a;@smith2006;@walters2000a;@white2011,;@white2010c]. In addition, even for the same total area of MPAs, the location and spacing of the MPAs can have a profound influence on their cumulative impact through network effects [@gaines2010; @costello2010a].

@@

The MPAs in this study cover 20% of the the surface waters of the Channel Islands, a region spanning roughly 90km east to west. Detailed dispersal studies are not available for all of the species covered by our study, but what information we have suggest that while adults of some of these species are likely to exhibit site fidelity [@love2011], larvae are likely distributed beyond the Channel Islands [@rassweiler2012]. Therefore, if we assume that the population of the targeted species is at least equal to the extent of the Channel Islands, we can assume that at most the MPAs cover roughly 20% of the targeted populations' ranges.

Formal stock assessments are largely lacking for these species. However, what evidence we have suggests mostly moderate fishing pressure, with some species such as California sheephead (*Semicossyphus pulcher*) and blue rockfish (*Sebastes mystinus*) experiencing high levels of fishing mortality / biomass  levels below target levels during the early 2000s [@alonzo2004;@dick2017].

We integrated these network design features, life history traits, and exploitation histories into a spatially explicit bio-economic simulation model, and used this model to generate expected outcomes for the Channel Islands MPAs, varying key unknowns such as the degree of pre-MPA depletion, movement rates, and fleet dynamics. While a wide range of outcomes are plausible, from 0% to upwards of 200%, the a median simulated effect size for a stylized version of the Channel Islands after over ten years of protection was roughly `r percent(median(sim_mpa_effects$mpa_effect[sim_mpa_effects$years_protected > 10]))`.

@@

Having established a theoretical grounding for the likely biomass density outcomes of the Channel Island MPAs, what empirical evidence do we have of the effects? @caselle2015 examined changes in biomass densities of species targeted by fishing activity inside and outside of the MPAs over time (a metric generally termed a "response ratio"), and compared these changes to the trends in non-targeted species. They found a statistically significant increase in the response ratios of targeted species over time, and evidence that this increase is smaller in the non-targeted species. Updating the results of @caselle2015 through 2017 with a Bayesian response-ratio regression shows a continuation of the increasing trend in the response ratios of targeted species (Fig.\@ref(fig:rr-plot)). This provides evidence that the Channel Islands MPAs are large enough and sufficiently well-enforced as to provide meaningful protection within their borders.

Why do these response ratio results not serve as sufficient evidence for the regional effects of the Channel Island MPAs? Control sites used in calculating response ratios are often selected based on abiotic or ecological traits such as habitat characteristics [@ferraro2018]. However, selection of control sites is further complicated by the very spillover that MPAs are often intended to create. Export of adults or larvae from the MPA to the "control" site affects their status as controls, as does displacement of fishing effort from MPAs to control sites. In theory, control sites far enough away to negate both biological spillover and concentration by the fishing fleet could be selected, but finding suitably far sites that are also appropriate proxies for the ecological and economic context of the MPAs may be challenging. As a result of these challenges, response ratios can be a highly imprecise and biased measure of the regional conservation effect of an MPA network [@ferraro2018].

<!-- While these concerns have been stated previously [e.g. @halpern2004;@white2011], the MPA evaluation literature has by and large been unable to adequately address them, often due to the very real logistical challenges of identifying and sampling adequate control sites that are truly independent of the MPAs in question -->

@@

While we do not have access to fine scale fishing data from the Channel Islands alone, reported catches for the species of interest in the Santa Barbara region in fact exhibit an overall downward trend in the years post reserve (see Fig.SXX). We can most likely rule out a negative MPA effect caused by a constant-catch fishing strategy then.  What is another possible explanation for the recent downward trend in the estimated MPA effects? The Channel Islands region (and the entire West coast of the USA) experienced a dramatic ‘marine heatwave’ beginning in 2014 and persisting through 2016, resulting in part in extremely elevated water temperatures throughout the region  [@gentemann2017]. Difference-in-difference models require a "parallel trends" assumption; in this case that in the absence of the MPAs targeted and non-targeted species would have exhibited similar trends in biomass densities. While this assumption cannot be proven, we find support for it in the initial years of the MPAs (Fig.SXX).
